---
title: "Getting Started with bivarCopula"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bivarCopula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette introduces the basic modeling workflow in `bivarCopula`.
The package is designed for Bayesian estimation of bivariate copula models
in which the marginal distributions and the dependence structure are
estimated jointly with Stan.

The statistical motivation follows directly from Sklar's theorem. Let
$X_1$ and $X_2$ denote two observed variables with marginal distribution
functions $F_1$ and $F_2$. Their joint distribution can be written as

$$
H(x_1, x_2) = C\{F_1(x_1), F_2(x_2); \theta\},
$$

where $C$ is a copula and $\theta$ is a dependence parameter. This
decomposition is useful because it separates two modeling decisions that
are often distinct in practice:

1. the choice of marginal distributions for each variable
2. the choice of a copula family for the dependence pattern

`bivarCopula` currently supports Gaussian, Clayton, and Joe copulas, in
combination with normal, lognormal, exponential, and beta marginals. The
package returns posterior draws for all unknown quantities, which allows
inference, uncertainty quantification, and model comparison within a
single framework.

## Prerequisites

`bivarCopula` depends on `cmdstanr` and therefore requires a working
CmdStan installation. The package website and README provide installation
instructions. Once the software stack is available, the vignette examples
begin with:

```{r}
library(bivarCopula)
library(copula)
```

The examples below simulate data with the `copula` package and then fit
the corresponding Bayesian model with `bivarCopula`.

## Model Specification in bivarCopula

The main entry point is `fit_bivariate_copula()`. Its first argument is a
numeric matrix with exactly two columns. Each row corresponds to one
observation, and each column corresponds to one variable.

```{r}
fit <- fit_bivariate_copula(
  U = data_matrix,
  copula = "gaussian",
  marginals = c("normal", "lognormal"),
  seed = 2024
)
```

Conceptually, model specification requires three inputs:

1. a two-column numeric matrix
2. a copula family
3. a pair of marginal distributions

The function returns a `copula_fit` object. This object stores the
underlying `CmdStanMCMC` fit together with metadata on the chosen copula,
marginals, and data dimensions. Standard accessors include `print()`,
`summary()`, and `coef()`.

### Data Requirements

The admissible support of the observed data depends on the chosen
marginals. The package checks these restrictions before sampling begins.

| Marginal | Parameterization | Data support |
|:---------|:-----------------|:-------------|
| `normal` | `mu`, `sigma` | Any real values |
| `lognormal` | `mu`, `sigma` on the log scale | Strictly positive |
| `exponential` | `lambda` | Strictly positive |
| `beta` | `alpha`, `beta` | Strictly between 0 and 1 |

If the original data are stored in a data frame, convert them to a matrix
before fitting:

```{r}
U <- as.matrix(df[, c("y1", "y2")])
```

### Available Copula Families

The supported copula families differ in the kind of dependence they can
represent.

| Copula | Parameter | Dependence pattern | Typical use |
|:-------|:----------|:-------------------|:------------|
| Gaussian | `rho` in `(-1, 1)` | Symmetric dependence, no tail dependence | General-purpose baseline model |
| Clayton | `theta_clayton > 0` | Lower-tail dependence | Joint occurrence of unusually small values |
| Joe | `theta_joe >= 1` | Upper-tail dependence | Joint occurrence of unusually large values |

The Gaussian copula is often a natural first specification because it is
simple to interpret and computationally stable. Clayton and Joe copulas
become attractive when theory or prior evidence suggests asymmetric tail
association.

### Available Marginal Distributions

The package currently supports four univariate marginal families.

| Marginal | Parameters | Remarks |
|:---------|:-----------|:--------|
| Normal | `mu`, `sigma` | Appropriate for approximately symmetric variables |
| Lognormal | `mu`, `sigma` | Appropriate for positive and right-skewed variables |
| Exponential | `lambda` | Appropriate for positive waiting-time type variables |
| Beta | `alpha`, `beta` | Appropriate for proportions in `(0, 1)` |

Different marginals can be assigned to the two variables, which is often
the main practical advantage of the copula approach.

## A Worked Example

This section illustrates a complete workflow for a Gaussian copula with a
normal first margin and a lognormal second margin.

### Step 1: Simulate Data

```{r}
seed <- 2024
set.seed(seed)

true_rho <- 0.5
n <- 2000

margins <- c("norm", "lnorm")
params <- list(
  list(mean = 0.8, sd = 2),
  list(meanlog = 0, sdlog = 0.8)
)

cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)
```

The simulated sample contains 2,000 observations from a joint model with
moderate positive dependence. In applied work, `data` would be replaced by
the observed two-column matrix of interest.

### Step 2: Fit the Model

```{r}
fit <- fit_bivariate_copula(
  data,
  copula = "gaussian",
  marginals = c("normal", "lognormal"),
  seed = seed
)
```

By default, the function runs four Markov chains, each with 1,000 warmup
iterations and 1,000 post-warmup iterations. These defaults are suitable
for many introductory analyses, although more demanding applications may
require longer runs.

### Step 3: Inspect the Posterior Summary

```{r}
print(fit)
summary(fit)
coef(fit)
```

These three accessors serve different purposes:

1. `print(fit)` provides a compact overview of the fitted model and a
   summary table for the model parameters.
2. `summary(fit)` returns the same information as a tibble, which is
   useful for programmatic inspection.
3. `coef(fit)` returns posterior means as a named numeric vector.

In this example, the posterior summary should recover the generating
parameter values with ordinary Monte Carlo variation. In particular, the
posterior mean of `rho` should be close to the true value of 0.5.

## Alternative Model Specifications

The same workflow applies to the remaining supported combinations of
copulas and marginals. The examples below illustrate several common
variants.

### Gaussian Copula with Normal and Exponential Marginals

```{r}
seed <- 2024
set.seed(seed)

true_rho <- 0.5
n <- 2000

margins <- c("norm", "exp")
params <- list(
  list(mean = 0.8, sd = 2),
  list(rate = 1)
)

cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(
  data,
  copula = "gaussian",
  marginals = c("normal", "exponential"),
  seed = seed
)

summary(fit)
```

This specification is useful when one variable is approximately symmetric
while the other is positive and strongly right-skewed.

### Gaussian Copula with Beta Marginals

```{r}
seed <- 2024
set.seed(seed)

true_rho <- 0.5
n <- 2000

margins <- c("beta", "beta")
params <- list(
  list(shape1 = 2, shape2 = 5),
  list(shape1 = 3, shape2 = 4)
)

cop <- normalCopula(param = true_rho, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(
  data,
  copula = "gaussian",
  marginals = c("beta", "beta"),
  seed = seed
)

summary(fit)
```

Beta marginals are appropriate for fractional outcomes or proportions that
lie strictly inside the unit interval.

### Clayton Copula with Normal and Lognormal Marginals

```{r}
seed <- 2024
set.seed(seed)

true_theta <- 2.0
n <- 2000

margins <- c("norm", "lnorm")
params <- list(
  list(mean = 0.8, sd = 2),
  list(meanlog = 0, sdlog = 0.8)
)

cop <- claytonCopula(param = true_theta, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(
  data,
  copula = "clayton",
  marginals = c("normal", "lognormal"),
  seed = seed
)

summary(fit)
```

The Clayton copula is especially relevant when joint lower-tail behavior
is substantively important. In financial or actuarial applications, this
can correspond to the simultaneous occurrence of unusually adverse
outcomes.

### Joe Copula with Normal and Lognormal Marginals

```{r}
seed <- 2024
set.seed(seed)

true_theta <- 2.0
n <- 2000

margins <- c("norm", "lnorm")
params <- list(
  list(mean = 0.8, sd = 2),
  list(meanlog = 0, sdlog = 0.8)
)

cop <- joeCopula(param = true_theta, dim = 2)
mvdc_copula <- mvdc(cop, margins = margins, paramMargins = params)
data <- rMvdc(n, mvdc_copula)

fit <- fit_bivariate_copula(
  data,
  copula = "joe",
  marginals = c("normal", "lognormal"),
  seed = seed
)

summary(fit)
```

The Joe copula is the corresponding choice when the focus is on the upper
tail, that is, when unusually large values tend to occur together more
often than would be implied by a Gaussian copula.

## Diagnostics and Practical Estimation Issues

Posterior inference is only meaningful when the Markov chains have mixed
well and the Hamiltonian Monte Carlo sampler has behaved adequately. The
most important diagnostics are available directly from the summary table
and from CmdStan's console output.

### Convergence Diagnostics

The `summary()` method reports `rhat`, `ess_bulk`, and `ess_tail` for each
parameter.

```{r}
summ <- summary(fit)

all(summ$rhat < 1.01)
all(summ$ess_bulk > 400)
all(summ$ess_tail > 400)
```

As a practical guideline:

1. `rhat` should usually be below 1.01 for every parameter.
2. `ess_bulk` and `ess_tail` should be large enough that Monte Carlo error
   is small relative to the posterior uncertainty.

If these conditions fail, increase the number of warmup or sampling
iterations and refit the model.

### Divergent Transitions

CmdStan reports divergent transitions after sampling. Divergences suggest
that the geometry of the posterior is difficult for the sampler and that
the numerical approximation may be unreliable.

```{r}
fit <- fit_bivariate_copula(
  data,
  copula = "gaussian",
  marginals = c("normal", "lognormal"),
  adapt_delta = 0.95,
  seed = seed
)
```

Increasing `adapt_delta` is often the first response. If divergences
persist even with values near 0.99, it is reasonable to investigate
whether the copula family or marginal assumptions are poorly aligned with
the observed data.

### Computational Controls

Several arguments can be adjusted to improve estimation:

| Argument | Role |
|:---------|:-----|
| `iter_sampling` | Number of post-warmup iterations per chain |
| `iter_warmup` | Number of warmup iterations per chain |
| `chains` | Number of Markov chains |
| `parallel_chains` | Number of chains run in parallel |
| `adapt_delta` | Target acceptance rate for NUTS |
| `max_treedepth` | Maximum tree depth for NUTS |

In substantive applications, it is good practice to document these
settings explicitly in any empirical report.

## Model Comparison with Leave-One-Out Cross-Validation

The Stan program stores the pointwise log-likelihood in `log_lik`. This
permits approximate leave-one-out cross-validation with the `loo` package.

```{r}
library(loo)

loo_result <- loo(fit)
print(loo_result)
```

To compare two candidate models, estimate both and then contrast their LOO
objects:

```{r}
fit_gaussian <- fit_bivariate_copula(
  data,
  copula = "gaussian",
  marginals = c("normal", "lognormal"),
  seed = seed
)

fit_clayton <- fit_bivariate_copula(
  data,
  copula = "clayton",
  marginals = c("normal", "lognormal"),
  seed = seed
)

loo_gaussian <- loo(fit_gaussian)
loo_clayton <- loo(fit_clayton)

loo_compare(loo_gaussian, loo_clayton)
```

The preferred model is the one with the larger expected log predictive
density, subject to the usual uncertainty reported by `loo`.

## Prior Specification

The Stan implementation uses weakly informative default priors. These
priors are intended to regularize estimation without imposing highly
restrictive substantive assumptions.

### Marginal Distribution Priors

| Distribution | Parameter | Prior | Interpretation |
|:-------------|:----------|:------|:---------------|
| Normal | `mu` | `Normal(0, 5)` | Diffuse location prior |
| Normal | `sigma` | `Lognormal(0, 1)` | Positive scale prior with median 1 |
| Lognormal | `mu` | `Normal(0, 5)` | Diffuse log-location prior |
| Lognormal | `sigma` | `Lognormal(0, 1)` | Positive log-scale prior with median 1 |
| Exponential | `lambda` | `Lognormal(0, 1)` | Positive rate prior with median 1 |
| Beta | `alpha` | `Gamma(2, 0.5)` | Positive shape prior with mean 4 |
| Beta | `beta` | `Gamma(2, 0.5)` | Positive shape prior with mean 4 |

### Copula Parameter Priors

| Copula | Parameter | Prior | Interpretation |
|:-------|:----------|:------|:---------------|
| Gaussian | `rho` | `Uniform(-1, 1)` | Flat prior over the admissible correlation range |
| Clayton | `theta_clayton` | `Lognormal(0, 1)` | Positive prior with median 1 |
| Joe | `theta_joe` | `Lognormal(log(2), 0.5)` | Prior centered on moderate upper-tail dependence |

The current R interface does not expose custom prior specification.
Researchers who require different priors can modify the Stan source file
at `inst/stan/fit_bivariate_copula.stan`.

## Mathematical Details

This section records the copula densities used by the package. In all
cases, the joint density for $(X_1, X_2)$ can be written as

$$
f(x_1, x_2) = c\{F_1(x_1), F_2(x_2); \theta\} f_1(x_1) f_2(x_2),
$$

where $c$ denotes the copula density and $f_1$, $f_2$ are the marginal
densities.

### Gaussian Copula

Define the probability integral transforms

$$
U_1 = F_1(X_1), \quad U_2 = F_2(X_2),
$$

and the corresponding normal scores

$$
Z_1 = \Phi^{-1}(U_1), \quad Z_2 = \Phi^{-1}(U_2).
$$

The Gaussian copula density is

$$
c(u_1, u_2; \rho) =
\frac{1}{\sqrt{1-\rho^2}} \exp\left(
  -\frac{1}{2(1-\rho^2)}(z_1^2 + z_2^2 - 2\rho z_1 z_2)
  + \frac{z_1^2}{2}
  + \frac{z_2^2}{2}
\right).
$$

Its log-density is

$$
\log c(u_1, u_2; \rho)
= -\frac{1}{2} \log(1-\rho^2)
+ \frac{\rho z_1 z_2 - \frac{1}{2}\rho^2(z_1^2 + z_2^2)}{1-\rho^2}.
$$

### Clayton Copula

The Clayton copula distribution function is

$$
C(u, v; \theta) =
\left(u^{-\theta} + v^{-\theta} - 1\right)^{-1/\theta},
\quad \theta > 0.
$$

Its density is

$$
c(u, v; \theta)
= (\theta + 1) u^{-\theta - 1} v^{-\theta - 1}
\left(u^{-\theta} + v^{-\theta} - 1\right)^{-1/\theta - 2}.
$$

The corresponding log-density is

$$
\log c(u, v; \theta)
= \log(1 + \theta)
+ (-\theta - 1)(\log u + \log v)
- \frac{2\theta + 1}{\theta}\log(u^{-\theta} + v^{-\theta} - 1).
$$

### Joe Copula

Write $\bar{u} = 1-u$ and $\bar{v} = 1-v$. The Joe copula distribution
function is

$$
C(u, v; \theta)
= 1 - \left[\bar{u}^{\theta} + \bar{v}^{\theta}
- \bar{u}^{\theta}\bar{v}^{\theta}\right]^{1/\theta},
\quad \theta \geq 1.
$$

Let $a = \bar{u}^{\theta}$, $b = \bar{v}^{\theta}$, and
$S = a + b - ab$. The density is

$$
c(u, v; \theta)
= \theta \bar{u}^{\theta-1} \bar{v}^{\theta-1}
S^{1/\theta - 2}
\left[
  S + \frac{\theta - 1}{\theta}(1-a)(1-b)
\right].
$$

The corresponding log-density is

$$
\log c(u, v; \theta)
= \log \theta
+ (\theta - 1)(\log \bar{u} + \log \bar{v})
+ \left(\frac{1}{\theta} - 2\right)\log S
+ \log\left[
  S + \frac{\theta - 1}{\theta}(1-a)(1-b)
\right].
$$
